---
title: "Data Cleaning"
author: "Sara Elfring"
date: "2024-12-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NHANES)
library(tidyverse)
library(ggplot2)
library(gtsummary)
```

## First Look

```{r}
# Load in NHANES data
data("NHANES")
```

```{r}
# Select Covariates and Summarize
NHANES |> 
  select(BPSysAve, Age, BMI, Gender, Alcohol12PlusYr, 
         AlcoholDay, AlcoholYear) |> 
  summary()
```

We start with `r nrow(NHANES)` samples.

## Remove samples missing BPSysAve

```{r}
sample = NHANES |> 
  select(BPSysAve, Age, BMI, Gender, Alcohol12PlusYr, 
         AlcoholDay, AlcoholYear) |> 
  filter(!is.na(BPSysAve))
summary(sample)
```

There are now `r nrow(sample)` samples.

## Remove samples younger than 18
```{r}
sample = sample |> 
  filter(Age >= 18)
```

Alcohol-related questions were only asked of participants over the age of 18. There are now `r nrow(sample)` samples.

## Remove Samples Missing Alcohol12PlusYr

```{r}
# Removing samples missing Alcohol12PlusYr 
sample = sample |> 
  filter(!is.na(Alcohol12PlusYr))
summary(sample)

# Save the samples missing Alcohol12PlusYr for Comparison
excluded = NHANES |> 
  select(BPSysAve, Age, BMI, Gender, Alcohol12PlusYr, 
         AlcoholDay, AlcoholYear) |> 
  filter(Age >= 18, !is.na(BPSysAve), is.na(Alcohol12PlusYr))
```

We have `r nrow(sample)` samples.

## Compare the samples missing and not missing Alcohol12PlusYr

```{r}
# Histogram comparison of BPSysAve
ggplot()+
  geom_histogram(data = sample, mapping = aes(x = BPSysAve), 
                 fill = "blue", alpha = 0.5, color = "black")+
  geom_histogram(data = excluded, mapping = aes(x = BPSysAve), 
                 fill = "red", alpha = 0.5, color = "black")
```

The distribution of BPSysAve is similar between those missing Alcohol12PlusYr and those not missing it.

## Remove Samples Missing BMI

```{r}
# Removing samples missing BMI 
sample = sample |> 
  filter(!is.na(BMI))
summary(sample)

# Save the samples missing BMI for Comparison
excluded = NHANES |> 
  select(BPSysAve, Age, BMI, Gender, Alcohol12PlusYr, 
         AlcoholDay, AlcoholYear) |> 
  filter(Age >= 18, !is.na(BPSysAve), !is.na(Alcohol12PlusYr), is.na(BMI))
```

We have `r nrow(sample)` samples.

```{r}
# Histogram comparison of BPSysAve
ggplot()+
  geom_histogram(data = sample, mapping = aes(x = BPSysAve), 
                 fill = "blue", alpha = 0.5, color = "black")+
  geom_histogram(data = excluded, mapping = aes(x = BPSysAve), 
                 fill = "red", alpha = 0.5, color = "black")
ggplot()+
  geom_histogram(data = excluded, mapping = aes(x = BPSysAve), 
                 fill = "red", alpha = 0.5, color = "black")
```

The BPSysAve values for samples missing BMI fall in the middle of the distribution of BPSysAve for samples not missing BMI.

## Remove Samples Missing AlcoholDay

```{r}
# Removing samples missing AlcoholDay 
sample = sample |> 
  filter(!is.na(AlcoholDay))
summary(sample)

# Save the samples missing AlcoholDay for Comparison
excluded = NHANES |> 
  select(BPSysAve, Age, BMI, Gender, Alcohol12PlusYr, 
         AlcoholDay, AlcoholYear) |> 
  filter(Age >= 18, !is.na(BPSysAve), !is.na(Alcohol12PlusYr), !is.na(BMI),
         is.na(AlcoholDay))
```

We have `r nrow(sample)` samples.

```{r}
# Histogram comparison of BPSysAve
ggplot()+
  geom_histogram(data = sample, mapping = aes(x = BPSysAve), 
                 fill = "blue", alpha = 0.5, color = "black")+
  geom_histogram(data = excluded, mapping = aes(x = BPSysAve), 
                 fill = "red", alpha = 0.5, color = "black")
```
The distribution of BPSysAve is similar now between those missing AlcoholDay and those not missing it.

## Remove Samples Missing AlcoholYear

```{r}
# Removing samples missing AlcoholYear 
sample = sample |> 
  filter(!is.na(AlcoholYear))
summary(sample)

```
There is only one observation missing AlcoholYear removed at this point.
We have `r nrow(sample)` samples.

## Remove outliers
We removed subject 2489 that has an average of 82 drinks in a day that they drink. We believe this is likely a data error as that is an extremely high number of drinks, let alone alcoholic drinks, to consume in one day.
```{r}
# Remove outlier with average of 82 drinks in a day
order(sample$AlcoholDay, decreasing = T)[1]

sample = sample |> 
  filter(AlcoholDay < 80)

```


## Summary of Our Sample

```{r}
# Number of samples
nrow(sample)

# Summarize continuous confounders
sample |> 
  summarize(mean(BPSysAve), sd(BPSysAve), mean(Age), sd(Age),
            mean(BMI), sd(BMI))
# Summarize continuous alcohol variables
sample |> 
  summarize(mean(AlcoholDay), sd(AlcoholDay),
            mean(AlcoholYear), sd(AlcoholYear))

# Summarize categorical variables
sample |> 
  summarize(prop_male = sum(Gender == 'male')/length(Gender), 
            prop_drink = 
              sum(Alcohol12PlusYr == 'Yes')/length(Alcohol12PlusYr))
```
# Making Table 1
```{r}
# Summary Statistics of all variables comparing missing to non-missing
NHANES |> 
  select(BPSysAve, Age, BMI, Gender, Alcohol12PlusYr, 
         AlcoholDay, AlcoholYear) |> 
  filter(Age >= 18) |> 
  mutate(missing = if_else(is.na(BPSysAve) | is.na(Alcohol12PlusYr) |
           is.na(BMI) | is.na(AlcoholDay) | is.na(AlcoholYear) | 
             AlcoholDay > 80,
           "Missing", "Sample")) |> 
  tbl_summary(
    by = missing,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                    all_categorical() ~ "{n} ({p}%)"),
    missing = "no",
    label = list(BPSysAve ~ "Combined Systolic Blood Pressure",
                 AlcoholDay ~ "Average Alcoholic Drinks in 1 Day",
                 AlcoholYear ~ "Days Drinking Alcohol in 1 Year",
                 Alcohol12PlusYr ~ 
                   "12 or more Alcoholic Drinks in 1 Year: Yes")
  )
```

